The Unicode Cookbook \linebreak for Linguists

### 2015-11-16 book/chapters/introduction.tex
minor revision intro

For example, the writing of tone diacritics in Yoruba is often proclaimed to be
the right way to write, although many users of Yoruba orthography seem to be
perfectly fine with leaving them out. As pointed out by the proponents of the
official rules, there are some homographs when leaving out the tone diacritics
\citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
represent) are normally full of homophones, which is normally not a problem at
all for speakers of the language. More importantly, writing is not just a purely
functional tool, but just as importantly it is a mechanism to signal social
affiliation. By showing that you \textit{know the rules} of expressing yourself
in writing, others will more easily accept you as a worthy participant in their
group. And that just as well holds for obeying to the official rules when
writing a job application, as for obeying to the informal rules when writing an
SMS to classmates in school. The case of Yoruba writing is an exemplary case, as
even after more than a century of efforts to standardize the writing systems,
there is still a wide range of variation in daily use \citep{Olumuyiw2013}.

The sometimes cumbersome and sometimes illogical structure, and the enormous
variability of existing writing systems is a fact of life scholars have to
accept and should try to adapt to as good as possible. Our plea here is a
proposal for a formalization to do exactly that.

When considering the worldwide linguistic diversity, including all
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format (a process called
\textsc{tailoring} in Unicode parlance). We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode locale descriptions, though we will argue that
in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of linguistics. Consequently, a central goal of
this paper is to flesh out the linguistic challenges for locale descriptions,
and work out suggestions to improve their structure for usage in a linguistic
context. Conversely, we also aim to improve linguists' understanding and
appreciation for the accomplishments of the Unicode Consortium in the
development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, in our
current research our goal is to develop quantitative methods for language
comparison and historical analysis in order to investigate worldwide linguistic
variation and to model the historical and areal processes that underlie
linguistic diversity,
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}.
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the \textit{same} language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme
separation of strings of symbols, a process we call \textsc{grapheme
tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
sentences) in a specific source, our first processing step is to specify how
these strings have to be separated into graphemes, considering the specific
orthographic conventions used in a particular source document. Our experience is
that such a graphemic tokenization can be performed without extensive in-depth
knowledge about the phonetic and phonological details of the language in
question. For example, the specification that $<$ou$>$ is a grapheme of English
is a much easier task than to specify what exactly the phonetic values of this
grapheme are in any specific occurrence in English words. Grapheme separation is
a task that can be performed relatively reliably and with limited availability
of time and resources (compare, for example, the task of creating a complete
phonetic or phonological normalization).

@@

Before we deal with these proposals, in the first part of this paper (Sections
~\ref{encoding} through~\ref{ipa-meets-unicode}) we give an extended introduction
to the notion of encoding (Section~\ref{encoding}) and writing systems, both
from a linguistic perspective and from the perspective of the Unicode Consortium
(Section~\ref{terminology}). We consider the Unicode Standard to be a
breakthrough (and ongoing) development that fundamentally changed the way we
look at writing systems, and we aim to provide here a slightly more in-depth
survey of the many techniques that are available in the standard. A good
appreciation for the solutions that the Unicode Standard also allows for a
thorough understanding of possible pitfalls that one might encounter when using
it (Section~\ref{unicode-pitfalls}). As an example of the current
state-of-the-art, we discuss the rather problematic marriage of the
International Phonetic Alphabet (IPA) with the Unicode Standard (Section
\ref{ipa-meets-unicode}).

The second part of the paper (Sections~\ref{orthography-profiles} and
\ref{use-cases}) describes our proposals for how to deal with the Unicode
Standard in the daily practice of (comparative) linguists. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Section~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Section~\ref{use-cases}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.

>>>>

For example, the writing of tone diacritics in Yoruba is often proclaimed to be
the right way to write, although many users of Yoruba orthography seem to be
perfectly fine with leaving them out. As pointed out by the proponents of the
official rules, there are some homographs when leaving out the tone diacritics
\citep[44]{Olumuyiw2013}. However, writing systems (and the languages they
represent) are normally full of homographs (and homophones), which is normally
not a problem at all for speakers of the language. More importantly, writing is
not just a purely functional tool, but just as importantly it is a mechanism to
signal social affiliation. By showing that you \textit{know the rules} of
expressing yourself in writing, others will more easily accept you as a worthy
participant in their group. And that just as well holds for obeying to the
official rules when writing a job application, as for obeying to the informal
rules when writing an SMS to classmates in school. The case of Yoruba writing is
an exemplary case, as even after more than a century of efforts to standardize
the writing systems, there is still a wide range of variation of writing in
daily use \citep{Olumuyiw2013}.

\subsubsection*{Formalizing orthographic structure}

The resulting cumbersome and often illogical structure of writing systems, and
the enormous variability of existing writing systems for the world's languages
is a fact of life scholars have to accept and should try to adapt to as good as
possible. Our plea in this book is a proposal how to do exactly that: formalize 
knowledge about individual writing systems in a form that is both easy to use 
for linguists in daily practice, and at the same time computer-readable for 
automatic processing.

When considering the worldwide linguistic diversity, including the many
lesser-studied and endangered languages, there exist numerous different
orthographies using symbols from the same scripts. For example, there are
hundreds of orthographies using Latin-based alphabetic scripts. All of these
orthographies use the same symbols, but these symbols differ in meaning and
usage throughout the various orthographies. To be able to computationally use
and compare different orthographies, we need a way to specify all orthographic
idiosyncrasies in a computer-readable format (a process called
\textsc{tailoring} in Unicode parlance). We call such specifications
\textsc{orthography profiles}. Ideally, these specifications have to be
integrated into so-called Unicode locale descriptions, though we will argue that
in practice this is often not the most useful solution for the kind of problems
arising in the daily practice of linguistics. Consequently, a central goal of
this book is to flesh out the linguistic challenges for locale descriptions,
and work out suggestions to improve their structure for usage in a linguistic
context. Conversely, we also aim to improve linguists' understanding and
appreciation for the accomplishments of the Unicode Consortium in the
development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, the proper
processing of orthographies and transcriptions becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity.\footnote{
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme
separation of strings of symbols, a process we call \textsc{grapheme
tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words,
sentences) in a specific source, our first processing step is to specify how
these strings have to be separated into graphemes, considering the specific
orthographic conventions used in a particular source document. Our experience is
that such a graphemic tokenization can often be performed reasonably accurately
without extensive in-depth knowledge about the phonetic and phonological details
of the language in question. For example, the specification that <ou> is a
grapheme of English is a much easier task than to specify what exactly the
phonetic values of this grapheme are in any specific occurrence in English
words. Grapheme separation is a task that can be performed relatively reliably
and with limited availability of time and resources (compare, for example, the
daunting task of creating a complete phonetic or phonological normalization).

@@

\subsubsection*{Structure of this bok}

Before we deal with these proposals we will extensively discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and writing
systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard also
allows for a thorough understanding of possible pitfalls that one might
encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
current state-of-the-art, we discuss the rather problematic marriage of the
International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
\ref{ipa-meets-unicode}).

The second part of the book (Sections~\ref{orthography-profiles} and
\ref{use-cases}) describes our proposals for how to deal with the Unicode
Standard in the daily practice of (comparative) linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Chapter~\ref{use-cases}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.
